---
title: "Toronto Raptors Shot Chart Report"
author: "Lucas Strauss"
date: "2025-04-03"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, include=FALSE}
library(nbastatR)
library(tidyverse)
library(ggplot2)
library(ggimage)
library(png)
library(ggpubr)
library(knitr)
Sys.setenv("VROOM_CONNECTION_SIZE"=131072*5)
```

```{r, echo=FALSE, warning=FALSE, include=FALSE}
#Creating and transforming dataframe
court <- readPNG("court.png") #Reading in background image
cool_colors <- c("#BA0C2F","black") #Changing theme colors to match Toronto Raptors colors
raptorsAll <- teams_shots(teams="Toronto Raptors", seasons = 2003) #First season defined directly to establish column names
for(x in 1:7) { #For loop used to add each season to one large dataframe (raptorsAll)
  raptors = teams_shots(teams="Toronto Raptors", seasons = (2003+3*x)) #Gather every third season for the dataframe
  raptorsAll <- bind_rows(raptorsAll, raptors)
}
  raptorsAll <- mutate(raptorsAll, isShotMade = as.factor(isShotMade)) %>% #Changing isShotMade to factor to change the legend
  mutate(isShotMade = recode(isShotMade, "TRUE"="Made", "FALSE"="Missed")) %>% #Change names so the legend will say Made or Missed instead
  mutate(isShotMade = factor(isShotMade, levels=c("Made", "Missed"))) #Establish levels so Made shots come first
```

```{r, fig.width=12, fig.height=7, echo=FALSE, warning=FALSE}
#Creating the shot chart
bball <- ggplot(data=raptorsAll, mapping = aes(x=locationX, y=locationY)) +
  background_image(court) + #Set background image to basketball court
  geom_point(aes(color=isShotMade), alpha=.2, size=.5, show.legend = TRUE) + #Create points for each shot separated into made and missed shots by color
  scale_x_continuous(limits=c(-250,250), breaks=NULL, expand=c(0,0)) + #Set the dimensions of baskeball court
  scale_y_continuous(limits=c(-40, 470), breaks=NULL, expand=c(0,0)) +
  labs(title = "Toronto Raptors Shot Charts", x="", y="", color="") + #Create title and get rid of axis and legend labels for clarity
  theme_minimal() +
  scale_color_manual(values=cool_colors) + #Set colors to Toronto Raptors colors
  facet_wrap(~yearSeason, nrow=2) + #Facet each shot chart by year
  guides(color = guide_legend(override.aes = list(size=3))) + #Increase the size of the dots in the legend
  theme(title = element_text(size=16), axis.text = element_blank(), legend.text = element_text(size=16), strip.text = element_text(size=14)) #Adjust sizes of labels for clarity
bball

```


- Shot charts were taken in three year intervals to demonstrate the evolution of the game

- Shot selection has changed tremendously over the past twenty years

- Midrange shots have virtually disappeared from outside the paint

- Three point shots and layups are far more common

- Three point shots from farther away are becoming more common

- The paint is utilized much more than in the past

- The baseline is rarely used anymore compared to the early 2000s

\newpage

## Detailed Report

- Shot charts were taken in three year intervals to demonstrate the evolution of the game
  - A range of twenty-two years is displayed
  - Red dots represent made shots while black dots represent missed shots
  - Each year has around 7000 data points so total density remains relatively consistent across years
- Shot selection has changed dramatically over the past twenty years
  - This is easily seen by the dot density changing locations over the years
  - The changes appear very gradually, which is why a range of twenty-two years is necessary
- Mid range shots have virtually disappeared from outside the paint
  - Shot density has moved away from mid range shots (i.e. just inside the three point line)
  - This makes sense because these shots have the lowest number of points per shot
- Three point shots and layups are far more common
  - This demonstrates smart decisions by teams by focusing more on easier shots or shots worth more
- Three point shots from farther away are becoming more common
  - This could be an effect of a stronger emphasis on three point shots
  - This could also be an effect of the NBA players attempting to be more stylistic to gain popularity
- The paint is utilized much more than in the past
  - Shot density shows a larger emphasis on paint play
  - This could be an effect of high skilled centers becoming more common
- The baseline is rarely used anymore compared to the early 2000s
  - Shots have transitioned away from the baseline
  - This makes sense because shots from near the baseline have a lower probability of makes

\newpage

## Four Point Shot

- As shots farther from the basket become more common, it is natural to think about a four point shot line
- To determine where the line should be placed, we must look at shot data across the NBA for two and three point shots from the last five years:


```{r, echo=FALSE, include=FALSE}
#Gleaning data for statistical and grpahical analysis of different shot types
numBins <- 30 #Set the number of bins for the chart later on
nbaLastFive <- teams_shots(all_active_teams = TRUE, seasons = 2020) #Initialize the dataframe with one year's data
for(x in 2021:2024) { #For loop used to add each season to the dataframe
  newYear = teams_shots(all_active_teams = TRUE, seasons = x)
  nbaLastFive <- bind_rows(nbaLastFive, newYear)
}

#Modify dataframe to include data needed for later analysis
nbaLastFive <- nbaLastFive %>%
  mutate(ShotDistance=sqrt(locationX*locationX+locationY*locationY)/10) %>% #Adding a column calculating distance from rim using Pythagorean Theorem
  mutate(Points = ifelse(typeShot=="2PT Field Goal", 2, 3)) %>% #Adding numerical value to type of shot
  mutate(bin = cut(ShotDistance, breaks=numBins, labels=FALSE)) #Separating shots into bins based on distance. This is needed to calculate the shot percentage at different distances
```


```{r, echo=FALSE}
#Creating a table to display meaningful statistics on two and three point shots
shotTable <- nbaLastFive %>%
  group_by(typeShot) %>% #Separating data into two and three point shots
  summarize(NumShots = n(), #Total number of shots
            PercentMade=sum(isShotMade==TRUE)/n(), #Field goal percentage
            PointsPerShot=mean(Points)*sum(isShotMade==TRUE)/n(), #Average number of points per shot
            AvgDistShot=mean(ShotDistance)) #Average shot distance from the basket
  kable(shotTable) #Making a polished graph to display
```


- This data demonstrates that the points per shot for two and three point shots are roughly the same (around 1.075)
- Thus, the four point shot should then also have a points per shot value around 1.075
- This would require the shot percentage to be around 27%
- To find a good distance to place the line, we can look at shot percentages at different distances


```{r, echo=FALSE}
#Creating dataframe of shot percentages at different distances
shotPercentBinned <- nbaLastFive %>%
  group_by(bin) %>% #Grouping by shot distance intervals created earlier
  summarize(shotAverage=mean(isShotMade)) %>% #Creating a shot average for each interval
  mutate(bin=bin*(max(nbaLastFive$ShotDistance)/numBins)) #Making the bin values correctly represent the distances by making it a fraction of the total number of bins and multiplying that by the maximum shot distance

#Creating the plot of field goal percentage over different distances
shotPercents <- ggplot(shotPercentBinned, mapping=aes(x=bin, y=shotAverage)) +
  geom_smooth(stat="identity", color="#BA0C2F") + #Line created without any statistical transformation to directly represent data
  labs(x="Shot Distance (ft)", y="Field Goal Percentage", title="Field Goal Percentage at Different Distances") +
  theme_minimal() +
  geom_hline(yintercept=.27) #Horizontal line representing the targeted 27% shot percentage
shotPercents
```


- This graph shows an intersection of 26% field goal percentage at around 34 feet
- However, after 34 ft, the field goal percentage drops sharply
- To average out the field goal percentage, we can move the line slightly closer to 32 ft
- The line would look like this:

```{r, echo=FALSE, message=FALSE}
#Creating points with distance of 32 ft away
fourPointLine <- data.frame(x=-250:250) %>% 
  mutate(y=sqrt((320*320)-(x*x))) #Using Pythagorean Theorem to place points with a distance of 32 ft

newCourt <- ggplot(fourPointLine, mapping=aes(x,y)) +
  background_image(court) + #Set background image to basketball court
  geom_smooth(color="black") +
  scale_x_continuous(limits=c(-250,250), breaks=NULL, expand=c(0,0)) + #Set the dimensions of baskeball court
  scale_y_continuous(limits=c(-40, 470), breaks=NULL, expand=c(0,0)) +
  labs(title = "", x="", y="") + #Create title and get rid of axis and legend labels for clarity
  theme_minimal()
newCourt
```


- This line would significantly shift the average distance of shots across the league
- If the points per shot is consistent as calculated, the total points per game should not change too much
- However, possessions might become much shorter, as quick four point shots become more common
